---
title: "Classification Tree - Vent-Flow Severity in Alberta Wells"
output:
  pdf_document: default
  word_document: default
  html_document: default
date: "2025-06-08"
---
```{r}
library(rpart)         
library(caret)        
library(dplyr)       
```

```{r}
data= read.csv("C:/Users/megha/Downloads/Gasreport_Cleaneddata.csv",stringsAsFactors = FALSE,check.names = FALSE)

data$Classification=as.factor(data$Classification)
data$Type =as.factor(data$Type)
data$Flow_Substance=as.factor(data$Flow_Substance)
dim(data)
```

```{r}
num_col=c("Flow_Rate_m3_day", "Stabilized_Shut_In_Pressure_kPa", 
             "Source_Depth_mkb", "Ground_Water_Base_mkb")
data=data[!apply(data[,num_col]==0,1,all),]
dim(data)
```

```{r}
table(data$Classification)
```
```{r}
table(data$Flow_Substance,data$Classification)
```



### Stratified K-Fold Cross-Validation : As classes are imbalanced, it ensures fair model evaluation
```{r}
set.seed(123)  
folds=createFolds(data$Classification, k = 10, list = TRUE, returnTrain = FALSE)
```

```{r}
results=data.frame(Accuracy = numeric(),MisclassError=numeric() ,Fold = integer())

for (i in 1:length(folds)) {
  test_idx=folds[[i]]
  test_data=data[test_idx, ]
  train_data=data[-test_idx, ]
  
  model=rpart(Classification ~ Flow_Rate_m3_day +
                Stabilized_Shut_In_Pressure_kPa+ 
                Source_Depth_mkb + Ground_Water_Base_mkb+Flow_Substance,
              data = train_data, method = "class")
  
  preds=predict(model, test_data, type = "class")#prediction
  #preds=factor(preds, levels =levels(test_data$Classification))
  actual=factor(test_data$Classification, levels =levels(train_data$Classification))
  acc=mean(preds == actual) #Accuracy  
  misclass=1 - acc #misclassification
  
  results=rbind(results, data.frame(Accuracy = acc, MisclassError = misclass,Fold= i))
}

print(results)
```
```{r}
table(test_data$Classification)
table(train_data$Classification)
```

```{r}
#Average accuracy : model's performance on test/unseen data
avg_acc=mean(results$Accuracy)
cat("Mean Accuracy:", round(avg_acc, 4))
```
```{r}
avg_missclass=mean(results$MisclassError)
cat("Mean Misclassification Error:",round(avg_missclass,4))
```
- Mean Misclassification Error: 0.1418, low misclassification error.

```{r}
train_preds=predict(model, train_data, type = "class")
train_acc=mean(train_preds == train_data$Classification)
cat("Training Accuracy:", round(train_acc, 4))
```
### No overfitting : training accuracy and cross validation accuracy is similar.

```{r}
conf_mat=confusionMatrix(preds, test_data$Classification)
conf_mat
```
- Accuracy : 0.8405, overall correctly predicted accuracy.
- Sensitivity : 0.9750, Correctly predict Non Serious.
- Specificity : 0.4519, Correctly predict Serious.
- Balanced Accuracy : 0.7135, Adjusts for class imbalance.
```{r}
library(ggplot2)
library(reshape2)
cm_df=as.data.frame(conf_mat$table)
ggplot(cm_df, aes(x = Prediction, y = Reference, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), vjust = 0.5, size = 5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal() +
  labs(title = "Confusion Matrix", x = "Predicted", y = "Actual")
```


```{r}
library(rpart.plot)
rpart.plot(model)
```
- Model used Flow_Rate_m3_day,Flow_Substance, Stabilized_Shut_In_Pressure_kPa, Source_Depth_mkb as key features.
- High flow rate is a strong indicator of Serious.
- High pressure increases the likelihood of a Serious prediction.
- Low pressure with deep sources : Serious.
- Shallow sources with low pressure : Non Serious.

"As class imbalance is there, there maybe be bias towards Non Serious (it has higher number of data)"

